Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-04 revised_main.tex
Updates from Overleaf

We provide a review of past works in order to understand what is lacking in the field, and how it can be improved. Through our review of the history of sound-matching, we provide a nomenclature for its different subtypes, and illuminate major perennial issues that prevent sound-matching as a practical approach for sound-designers. Here we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, we find that not much is known about the interaction between different methods of synthesis and different loss functions. While there have been many works and experiments comparing the accuracy of loss functions and choosing the best performers~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that these claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings as typically only a single (and often, simple) synthesizer has been used for testing the effectiveness of loss functions. Should we be searching for a universally best performing loss function across synthesizers? Does the method of synthesis influence the selection of the ``optimal" loss function, and if so, do current, SOTA loss functions satisfy the creative needs of sound-designers?

The main hypothesis here is that \textit{the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis}. For the experiments, a \textit{differentiable}, \textit{iterative} approach to sound-matching is used. The iterative approach  better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while allowing access to the gradients that can be used to better understand the nature of the problem.

% The simple synthesizers tested here represent fundimental building blocks of more complex implementations, and direct use of loss function gradients assuming

We combine 4 loss functions with 4 differentiable synthesizers, and compare their performance as measured by the similarity of the final output to the target. We evaluate the final similarity with two different automatic methods, as well as manual listening tests conducted by two of the authors. We compare the distributions of the evaluation scores in order to rank the loss functions from best to worst. The results of these experiments show that when accounting for different synthesizers, there is no consistent choice of sound similarity measure for optimal performance in sound-matching.

Here, several common characteristics of previous work are avoided in order to gain fundamental, practical takeaways from the experiments. In order to directly analyze the interaction between the loss and method of synthesis, the synthesizer here are implemented with simple, functions that are used as the building blocks of complex synthesizers, and the gradients flow from the loss function to the synthesizer parameters with no extra layers between the two. Furthermore, since the true target parameters are unknown in any realistic sound-design scenario, parameter loss (P-Loss) is only used as a secondary measure of performance, and not used as a loss function.

%   A primary conclusion of the work is that there is no one best sound similarity measure capable of working across all synthesizers.
% The core problem that no single loss function performs optimally across synthesizer types

>>>>

We provide a review of past works in order to understand what is lacking in the field, and how it can be improved. Through our review of the history of sound-matching, we provide a nomenclature for its different subtypes, and illuminate major perennial issues that prevent sound-matching as a practical approach for sound-designers. Here we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, we find that not much is known about the interaction between different methods of synthesis and different loss functions. We believe that clarifying the relationship between these two components is a necessary step for the prioritization of future works towards practical sound-matching.

The main hypothesis here is that \textit{the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis}. While there have been many works and experiments comparing the accuracy of loss functions~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings; For example, it is typical for only a single (and often, simple) synthesizer to be used for testing the effectiveness of loss functions. This main hypothesis would help with important questions regarding the path of future research: Should we be searching for a universally best performing loss function across synthesizers? Does the method of synthesis influence the selection of the ``optimal" loss function, and if so, do current, SOTA loss functions satisfy the creative needs of sound-designers?

For the experiments, an \textit{iterative} and \textit{differentiable}, approach to sound-matching is used. The iterative approach better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while a differentiable environment allows access to the loss function gradients that can be used to better understand the nature of the problem. We define four differentiable synthesizers (each showcasing a fundamental method of synthesis used in modern synthesizers) and pair them with four different loss functions (two established and two novel methods). We evaluate the final similarity with two different automatic methods, as well as manual listening tests conducted by two of the authors. We compare the distributions of the evaluation scores in order to rank the loss functions from best to worst. The results of these experiments show that when accounting for different synthesizers, there is no consistent choice of sound similarity measure for optimal performance in sound-matching.

% Here, several common characteristics of previous work are avoided in order to gain fundamental, practical takeaways from the experiments. In order to directly analyze the interaction between the loss and method of synthesis, the synthesizer here are implemented with simple, functions that are used as the building blocks of complex synthesizers, and the gradients flow from the loss function to the synthesizer parameters with no extra layers between the two. Furthermore, since the true target parameters are unknown in any realistic sound-design scenario, parameter loss (P-Loss) is only used as a secondary measure of performance, and not used as a loss function.

% The simple synthesizers tested here represent fundimental building blocks of more complex implementations, and direct use of loss function gradients assuming 
%   A primary conclusion of the work is that there is no one best sound similarity measure capable of working across all synthesizers.
% The core problem that no single loss function performs optimally across synthesizer types
